---
title: "Statistical Inference Project"
author: "Shambhavi Malik"
date: "31/08/2020"
output: 
  html_document: default
  pdf_document: default
  md_document: default
---
## Simulation Exercise 
Overview

In this we investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. We investigate the distribution of averages of 40 exponentials. Note that we do a thousand simulations.

### Part 1: Show the sample mean and compare it to the theoretical mean distribution
```{r}
n <- 40
Simulations <- 1000
Lambda <- 0.2

SampleMean <- NULL
for(i in 1:Simulations) {
  SampleMean <- c(SampleMean, mean(rexp(n, Lambda)))
}
mean(SampleMean)
```

### Part 2: Show the sample is (via variance) and compare it to the thoretical variance of the distribtution
```{r}
Variance <- var(SampleMean)
Variance
```

### Part 3: Show that the distribution is appoximately normal
```{r}
hist(SampleMean, breaks = n, prob = T, col = "turquoise", xlab = "Means")
x <- seq(min(SampleMean), max(SampleMean), length = 100)
lines(x, dnorm(x, mean = 1/Lambda, sd = (1/Lambda/sqrt(n))), pch = 25, col = "black")
```

```{r}
qqnorm(SampleMean)
qqline(SampleMean, col = "red")
```

The distribution averages of 40 exponentials is very close to a normal distribution.

## Basic Inferential Data Ananlysis
We will analyze the ToothGrowth data in the R datasets package.

1. Load the ToothGrowth data and perform some basic exploratory data analysis
```{r}
library(datasets)
data(ToothGrowth)
library(ggplot2)

str(ToothGrowth)
```

2. Look at the data set
```{r}
head(ToothGrowth)
summary(ToothGrowth)
```

3. Plotting the data
```{r}
ggplot(data=ToothGrowth, aes(x=as.factor(dose), y=len, fill=supp)) +
    geom_bar(stat="identity") +
    facet_grid(. ~ supp) +
    xlab("Dose(mg)") +
    ylab("Tooth length")
```

4. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.
```{r}
hypoth1 <- t.test(len ~ supp, data = ToothGrowth)
hypoth1$conf.int
hypoth1$p.value
hypoth2<-t.test(len ~ supp, data = subset(ToothGrowth, dose == 0.5))
hypoth2$conf.int
hypoth2$p.value
hypoth3<-t.test(len ~ supp, data = subset(ToothGrowth, dose == 1))
hypoth3$conf.int
hypoth3$p.value
hypoth4<-t.test(len ~ supp, data = subset(ToothGrowth, dose == 2))
hypoth4$conf.int
hypoth4$p.value
```

### Conclusion
OJ ensures more tooth growth than VC for dosages 0.5 & 1.0. OJ and VC gives the same amount of tooth growth for dose amount 2.0 mg/day. For the entire trail we cannot conclude OJ is more effective that VC for all scenarios.