---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The zip file is in the current working directory which was cloned from GitHub. We first need to unzip the file to access its contents.
After unzipping we read the table activity.csv and store it in the variable data_activity.
See its first few entries using the head command.
```{r}
unzip("activity.zip")
data_activity <- read.csv("activity.csv")
head(data_activity)
```


## What is mean total number of steps taken per day?

We first form a data set of total steps taken per day using the aggregate function. We ignore the NA values.
View the first few contents of total_step using the head command.
```{r}
total_step<-aggregate(steps~date,data_activity,sum,na.rm=TRUE)
head(total_step)
```

Now we visualize the data by plotting a histogram.
```{r}
hist(total_step$steps,breaks= 20,col = "turquoise",xlab = "Total Steps per day",ylab = "Frequency")
```

We can see the mean of the dataset by using the summary command.
We can also calculate mean and median by using their respective commands.
```{r}
summary(total_step)
mean_steps <- mean(total_step$steps)
median_steps <- median(total_step$steps)
mean_steps
median_steps
```



## What is the average daily activity pattern?
We create a data set of steps according to 5 min intervals and take their average using the aggregate function.
Look at its contents using head command.
```{r}
meanStepsInterval <- aggregate(steps ~ interval, data_activity, mean)
head(meanStepsInterval)
```

Then we make a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
plot(x=meanStepsInterval$interval, y=meanStepsInterval$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval",
     ylab="Number of Steps", xlab="Intervals (in 5 mins)",
     col="darkblue", lwd=1.5)
```

To find the 5-minute interval for which, on average across all the days in the dataset, contains the maximum number of steps
```{r}
meanStepsInterval[grep(max(meanStepsInterval$steps), meanStepsInterval$steps), ]
```

## Imputing missing values

First we need to calculate total number of missing values in the dataset.
```{r}
sum(is.na(data_activity$steps))
sum(is.na(data_activity$date))
sum(is.na(data_activity$interval))
```

Now we replace the missing values by the mean steps in intervals and create a new data set.
```{r}
imputedData <- data_activity
for(x in 1:17568) {
    if(is.na(imputedData[x, 1])==TRUE) {
        imputedData[x, 1] <- meanStepsInterval[meanStepsInterval$interval %in% imputedData[x, 3], 2]
    }
}
head(imputedData)
```

```{r}
imputedTotalStepsDay <- aggregate(steps ~ date, imputedData, sum)
head(imputedTotalStepsDay)
```

Now we create a histogram of the new dataset.
```{r}
hist(imputedTotalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     main="Histogram of Total Number of Steps Taken per Day (With Imputed Values)",
     col="coral")
```

We can see the mean of the new dataset by using the summary command.
We can also calculate mean and median by using their respective commands.
```{r}
summary(imputedTotalStepsDay)
mean_steps_new <- mean(imputedTotalStepsDay$steps)
median_steps_new <- median(imputedTotalStepsDay$steps)
mean_steps_new
median_steps_new
```

Although the results are quite similar the total steps per day increases by imputing the data.
We can compare by plotting graphs of both data set side by side and using the same graph limits.
```{r}
par(mfrow = c(1, 2))
hist(total_step$steps, breaks=20, xlab="Number of Steps Taken", 
     col="turquoise", ylim=c(0, 20), main=NULL)
hist(imputedTotalStepsDay$steps, breaks=20, xlab="Number of Steps Taken", 
     col="coral", ylim=c(0, 20), main=NULL)
```

## Are there differences in activity patterns between weekdays and weekends?

We create a new data set by classifying days into weekdays and weekends.
```{r}
daysData <- imputedData
daysData$days <- weekdays(as.Date(daysData$date))
daysData$weekday <- as.character(rep(0, times=17568))
for(x in 1:17568) {
    if(daysData[x, 4] %in% c("Saturday", "Sunday")) {
        daysData[x, 5] <- "weekend"
    } else {
        daysData[x, 5] <- "weekday"
    }
}
daysData$weekday <- factor(daysData$weekday)
head(daysData)
```

Now we separate the data into 2 data sets according to weekdays.
```{r}
weekdayData <- daysData[daysData$weekday=="weekday", ]
weekendData <- daysData[daysData$weekday=="weekend", ]
weekdayMean <- aggregate(steps ~ interval, weekdayData, mean)
weekendMean <- aggregate(steps ~ interval, weekendData, mean)
head(weekdayMean)
head(weekendMean)
```

Finally we plot for both the data sets.
```{r}
par(mfrow=c(2, 1),mar=c(4, 4, 3, 2))
plot(weekdayMean$interval, weekdayMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekdays",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps",
     col="steelblue", lwd=1.5, ylim=c(0, 230))
plot(weekendMean$interval, weekendMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekends",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps",
     col="springgreen", lwd=1.5, ylim=c(0, 230))
```